Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-24 _paper_output/paper_sparse_hawkes.tex
mysterious declaration randomisation
cites

The Hawkes process (Hawkes, 1971) specifically, is the linear,
self-exciting point process; that is, given a "background" rate
$\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,$ and a non-negative influence
kernel with positive support
$\phi:\mathbb{R}^+\rightarrow\mathbb{R}^+\cap\{0\}$, the Hawkes process
has the following expression for its intensity:

@@

    One such assumption would be penalization :citeTikhonov1965,Hoerl1970
-{}-I apply an additional penalty to the log likelihood function,
penalizing particular values of the parameters.

For the penalized log likelihood for parameter $\theta$ and penalty
$\pi$, I write

@@

Hereafter, I will drop the prime from the augmented parameter vector and
simply write $\theta$.

Penalization is more frequently presented in the context of generalized
linear regression from i.id. samples, but it also fits within a
generalized Maximum Likelihood estimation theory.

Various gradient ascent algorithms are known to perform well for
:cite::Simon2011,Wu2008. Pragmatically, for performance reasons, I used
a mixed strategy. My algorithm attempts to update marginal estimates for
each $\omega_i$ parameter more rapidly first via Newton's method
:citeBattiti1992,Ozaki1979 and uses conjugate gradient descent for the
parametric estimates of the Hawkes parameters. If the updates steps are
small this seems to be stable. There are various tuning parameters to
such algorithms.

Many variants of pathwise regularization and exist, such as backwards
stepwise regularization, versions with random initialization, Least
Angle Regression, :citeEfron2004a selection with integrated cross
validation and so on. :citeFriedman2010

Many variants of pathwise regularization and exist, such as backwards
stepwise regularization, versions with random initialization, Least
Angle Regression, :citeEfron2004a selection with integrated cross
validation and so on. :citeFriedman2010 For this particular model I have
found few results giving me analytic guarantees, so I use heuristics and
simulation-based verification. I do not declare that this method "best"
in any sense, or that it will find the correct global maximum penalized
likelihood etc. The method, however, is simple enough to prototype
rapidly and, as it turns out, performs well on test data. Results will
be presented in the next chapter.

@@

Similarly, we will approximate $\phi$ in a different functional basis
${f}_{1\leq k \leq K}$, such that $\phi(s):= \sum_{k=1}^K f(s;k)$. This
choice is likely arbitrary, but many exist in the literature trading off
computational and statistical convenience, including Laguerre
polynomials (Ogata and Akaike, 1982), Exponential functions (Schoenberg
2015; Rambaldi et al, 2015), and step functions (Eichler, 2016). As a
convenient choice, we propose a basis of smooth unimodal non-negative
functions with a computationally convenient form and a single mode, the
convenient Maxwell distribution,
\[\phi_j(t):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma_j^2}\right)}{\sigma_j^3}.\]

@@

There is no closed-form solution for the optimum, but the derivatives of
this function are easy to calculate manually see e.g. (Ozaki 1979). Here
howevfer, we use automatic differentiation to make more complex
penalisation procedure conveneient to differentiate (Baydin and
Pearlmutter 2014).

@@

    To remedy this, I use Sugiura's finite-sample-corrected version, the
\emph{AICc} (Sugiura, 1978).
\[{\mathrm {AIC}(L(X;\hat{\theta}))}=2d-2\ln(L)\]

@@

The traditional means of performing model selection in Hawkes-type
models e.g. Ogata (1983), uses \emph{degrees-of-freedom} penalties
(Efron 1986). Specifically, the \emph{AIC}, the Akaike Information
Criterion (Akaike 1981) is widely popular.

@@

However, the AIC does not immediately solve the problem of
regularizition parameter selection in maximum likelihood, where the
regularisation parameter may be a real value smoothly interpolating many
possible models. One can calculate an AIC for penalised models, but this
require onerous Hessian matrix inversions and becomes costly in high
dimensions, and in any case does not generalise readily to non-smooth
penalties. (Konishi and kitagawa 2008)

We proceed heuristically, based on the arguments from Lasso-type
linear-model regularisation. There, the number of non-zero coefficients
fit under a particular value of the regularization penalty $\pi$ in an
$\ell_1$ regression model provides an unbiassed estimate of the
effective degrees of freedom of the model (Zou et al, 2007.)

@@

(Snoek et al, 2012) using the package \texttt{scikit-optimize} (andreh7
et al, 2017) .

@@

        <ipython-input-8-3150368f60c1> in <module>()
          1 from skopt import load
          2 
    ----> 3 res\_gp = load(join(FIT\_DIR, 'additive\_cvge'))

>>>>

The Hawkes process \cite{hawkes_point_1971} specifically, is the linear,
self-exciting point process; that is, given a "background" rate
$\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,$ and a non-negative influence
kernel with positive support
$\phi:\mathbb{R}^+\rightarrow\mathbb{R}^+\cap\{0\}$, the Hawkes process
has the following expression for its intensity:

@@

    One such assumption would be penalization -{}-I apply an additional term
to the log likelihood function, penalizing particular values of the
parameters. For the penalized log likelihood for parameter $\theta$ and
penalty $\pi$, I write

@@

For this particular model I have found few results giving me analytic
guarantees, so I use heuristics and simulation-based verification. I do
not declare that this method "best" in any sense, or that it will find
the correct global maximum penalized likelihood etc. The method,
however, is simple enough to prototype rapidly and, as it turns out,
performs well on test data. Results will be presented in the next
chapter.

@@

Similarly, we will approximate $\phi$ in a different functional basis
${f}_{1\leq k \leq K}$, such that $\phi(s):= \sum_{k=1}^K f(s;k)$. This
choice is likely arbitrary, but many exist in the literature trading off
computational and statistical convenience, including Laguerre
polynomials \cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_conistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form and a single mode, the convenient
Maxwell distribution,
\[\phi_j(t):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma_j^2}\right)}{\sigma_j^3}.\]

@@

There is no closed-form solution for the optimum, but the derivatives of
this function are easy to calculate manually see e.g.
\cite{ozaki_maximum_1979}. Here howevfer, we use automatic
differentiation to make more complex penalisation procedure conveneient
to differentiate \cite{baydin_automatic_2014}.

@@

    To remedy this, I use Sugiura's finite-sample-corrected version, the
\emph{AICc} \cite{sugiura_further_1978}.
\[{\mathrm {AIC}(L(X;\hat{\theta}))}=2d-2\ln(L)\]

@@

The traditional means of performing model selection in Hawkes-type
models e.g. \cite{ogata_estimation_1983}, uses \emph{degrees-of-freedom}
penalties \cite{efron_how_1886}. Specifically, the \emph{AIC}, the
Akaike Information Criterion \cite{akaike_likelihood_1981} is widely
popular.

@@

However, the AIC does not immediately solve the problem of
regularizition parameter selection in maximum likelihood, where the
regularisation parameter may be a real value smoothly interpolating many
possible models. One can calculate an AIC for penalised models, but this
require onerous Hessian matrix inversions and becomes costly in high
dimensions, and in any case does not generalise readily to non-smooth
penalties. \cite{konishi_information_2008}

We proceed heuristically, based on the arguments from Lasso-type
linear-model regularisation. There, the number of non-zero coefficients
fit under a particular value of the regularization penalty $\pi$ in an
$\ell_1$ regression model provides an unbiassed estimate of the
effective degrees of freedom of the model \cite{zou_degrees_2007}

@@

\cite{snoek_practical_2012} using the package \texttt{scikit-optimize}
\cite{andreh7_scikit-optimize_2017} .

@@

        <ipython-input-12-3150368f60c1> in <module>()
          1 from skopt import load
          2 
    ----> 3 res\_gp = load(join(FIT\_DIR, 'additive\_cvge'))
